\chapter*{Acknowledgments}

First and foremost I would like to thank my adviser professor Peter M{\"u}ller for his guidance, support, patience and trust. Peter has allowed me an almost free hand at exploring ideas while gently ensuring I do not diverge, and giving me all the help and support one could hope for.

I would like to thank my co-examiners, Professor Clark Barrett, Professor Arie Gurfinkel and Professor Martin Vechev for taking the time to review this long thesis.

I would also like to thank:\\
Milos Novacek for his support on and off the campus and for endless discussions.\\
Dr. Alexander Summers for his many many useful comments and for taking time for lengthy discussions.\\
Dimitar Asenov for his help with coding, diagrams, logistics and in general.\\
Lucas Brutchy and Malte Schwerhof for many useful discussions and suggestions.\\
Dr. Caterina Urban for her help and support in the final stages of the PhD.

I would like to thank Prof. Arie Gurfinkel for deep and lengthy discussions way past midnight and for his patience in guiding me towards relevant ideas and papers.

I would like to thank Dr. Leonardo de Moura for exposing me to the details and secrets of SMT solving.

A special thanks goes to Marlies Weissert for her invaluable help innumerable times in all manner of problems, on and off the campus.

The Programming Methodology group is a great place to work and explore ideas (and a great place in general) and I was fortunate to share it together with Dr. Hermann Lehner, Dr. Joseph Rusikewich, Dr. Pietro Ferrara, Dr. Arsenii Rudich, Dr. Valentin W{\"u}stholz, Dr. Maria Christakis, Dr. Yannis Kassios, Dr. C\'{e}dric Favre, Arshavir Ter-Gabrielyan, Marco Eilers, Alexandra Bugariu and J\'{e}r\^{o}me Dohrau.

I would also like to thank my good friend Sabrina Bahnm{\"u}ller for her support throughout the PhD period.

I would like to thank my parents Arie and Esther and my brother Amnon for their patience and support throughout the entire PhD period.




\newpage
